---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# security

<!-- badges: start -->
[![R-CMD-check](https://github.com/jabenninghoff/security/workflows/R-CMD-check/badge.svg)](https://github.com/jabenninghoff/security/actions)
[![lint](https://github.com/jabenninghoff/security/workflows/lint/badge.svg)](https://github.com/jabenninghoff/security/actions)
<!-- badges: end -->

## Overview

**`r desc::desc_get_field("Title")`:** `r desc::desc_get_field("Description")`

Feel free to use and/or fork this project!

## Installation

You can install the development version of security from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("jabenninghoff/security")
```

Or using renv:

``` r
# install.packages("renv")
renv::install("jabenninghoff/security")
```

## Development

- Changelog: See "Changelog" or `NEWS.md`.
- Planned: [TODO](TODO.md)

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
# basic example code
```
