---
title: "Constraints vs Performance"
author: "John Benninghoff"
date: '2024-04-30'
date-modified: '2024-05-10'
categories: notes
order: 107
output:
  html_notebook:
    theme:
      version: 5
      preset: bootstrap
    css: assets/extra.css
    pandoc_args: --shift-heading-level-by=1
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

Visualizations exploring the use of constraints vs performance improvements in risk management.

# Questions/TODO

- [ ] Questions/TODO list here

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(jbplot)
library(fs)
library(tibble)
library(dplyr)
```

# Normal Performance

Replicate a version of Figure 9 from the
[Safety-II White Paper](https://www.england.nhs.uk/signuptosafety/wp-content/uploads/sites/16/2015/10/safety-1-safety-2-whte-papr.pdf),
with help from <https://ggplot2tutor.com/tutorials/sampling_distributions>:

```{r background}
xmin <- -5
xmax <- 5

save_png <- function(filename) {
  ggsave(filename = path("rendered", filename), width = 16 * 0.6, height = 9 * 0.6, bg = "white")
}

background <- ggplot(data.frame(x = c(xmin, xmax)), aes(x)) +
  scale_x_continuous(breaks = -3:3, minor_breaks = NULL) +
  labs(x = NULL, y = NULL) +
  theme_quo(minor.y = FALSE)

baseline <- stat_function(fun = dnorm, geom = "line")
bad <- stat_function(fun = dnorm, geom = "area", fill = "red", xlim = c(xmin, -2))

background + bad + baseline

save_png("01-baseline-bad.png")
```

The plot above shows "bad" outcomes in red. Let's add in "good" outcomes (>1) in green:

```{r good}
good <- stat_function(fun = dnorm, geom = "area", fill = "green", xlim = c(1, xmax))

background + bad + good + baseline

save_png("02-baseline-bad-good.png")
```

# Constrained Performance

One way of reducing "bad" outcomes is by constraining performance - reducing the standard
deviation.

```{r constrained}
constrained <- stat_function(fun = dnorm, args = list(sd = 0.7), geom = "line", color = "blue")
taller <- scale_y_continuous(limits = c(0, 0.6))

background +
  stat_function(
    fun = dnorm, args = list(sd = 0.7), geom = "area", fill = "red", xlim = c(xmin, -2)
  ) +
  stat_function(
    fun = dnorm, args = list(sd = 0.7), geom = "area", fill = "green", xlim = c(1, xmax)
  ) +
  constrained +
  taller

save_png("03-constrained.png")
```

Plotting both on the same grid shows the reduction in both "bad" and "good" outcomes:

```{r overlay}
background +
  bad +
  good +
  baseline +
  constrained +
  taller

save_png("04-baseline-constrained.png")
```

# Improved Performance

Another way of reducing bad outcomes is by improving performance - shifting the mean.

```{r performance}
performance <- stat_function(fun = dnorm, args = list(mean = 1), geom = "line", color = "blue")
improved <- stat_function(
  fun = dnorm, args = list(mean = 1), geom = "area", fill = "green", xlim = c(1, xmax)
)

background +
  stat_function(
    fun = dnorm, args = list(mean = 1), geom = "area", fill = "red", xlim = c(xmin, -2)
  ) +
  improved +
  performance

save_png("05-improved.png")
```

Plotting both together shows a reduction in "bad" and an increase in "good" outcomes:

```{r overlay_perf}
background +
  bad +
  improved +
  baseline +
  performance

save_png("06-baseline-improved.png")
```

Comparing all three:

```{r overlay_all}
background +
  bad +
  improved +
  baseline +
  constrained +
  performance +
  taller

save_png("07-baseline-constrained-improved.png")
```

# Growth of Controls

Visualize an example of the growth of controls using the Cyentia/RiskRecon State of Third-Party Risk
Management 2020 and 2024 reports (data from 2023).

Source:

- 2020 data: <https://www.riskrecon.com/state-of-third-party-risk-management-report>
- 2023 data: <https://www.riskrecon.com/state-of-third-party-risk-management-2024>

```{r questionnaire}
questionnaire <- tribble(
  ~year, ~questions, ~percent,
  2020,  ">400",     0.041,
  2020,  "101-400",  0.148,
  2020,  "11-100",   0.705,
  2020,  "1-10",     0.107,
  2023,  ">400",     0.02,
  2023,  "101-400",  0.333,
  2023,  "11-100",   0.616,
  2023,  "1-10",     0.030
) |>
  mutate(year = as.factor(year)) |>
  mutate(questions = factor(questions, levels = c("1-10", "11-100", "101-400", ">400")))

ggplot(questionnaire, aes(questions, percent, fill = year)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("steelblue2", "steelblue4")) +
  labs(x = NULL, y = NULL, fill = "Year", title = "Third-Party Questionnaire Length by Year") +
  labs(caption = "Source: Cyentia/RiskRecon State of Third-Party Risk Management, 2020 and 2024") +
  theme_quo()

save_png("08-questionnaire-length.png")
```

# Transparent Donut

Create a transparent donut plot showing an 80% reduction.

```{r custom_donut}
# custom function based on ggplot_donut_percent()
custom_donut <- function(p, text = "", accuracy = NULL, hsize = 4, size = 12, family = "Lato") {
  data <- data.frame(group = c(TRUE, FALSE), n = c(p, 1 - p))
  label <- paste0(scales::label_percent(accuracy = accuracy)(p), "\n", text)

  ggplot_donut(data, hsize = hsize) +
    guides(fill = "none") +
    geom_text(x = 0, label = label, size = size, family = family) +
    scale_fill_grey() +
    theme(plot.background = element_rect(fill = "transparent", color = NA))
}

custom_donut(0.8, "reduction")
ggsave("rendered/80-percent-safety.png")

custom_donut(0.8, "reduction?")
ggsave("rendered/80-percent-security.png")
```
