---
title: "MFA Deployment Analysis"
author: "John Benninghoff"
date: '2026-09-11'
date-modified: '2026-09-11'
categories: [risk quantification, security differently]
order: 108
output:
  html_notebook:
    theme:
      version: 5
      preset: bootstrap
    css: assets/extra.css
    pandoc_args: --shift-heading-level-by=1
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

Analysis of the impact of Multi-Factor Authentication (MFA) deployment rate on the likelihood of compromise.

# Questions/TODO

- [ ] Questions/TODO list here

```{r}
#| label: setup
#| message: false
#| warning: false
library(purrr)
library(formattable)
library(ggplot2)
library(plotly)
library(jbplot)
```

# Background

Daniel Woods' research on the empirical effectiveness of security controls shows that Multi-Factor
Authentication (MFA) appears to reduce the likelihood of compromise by as much as 99% for
individuals, but implementing MFA doesn't show the same reduction for organizations, discussed in
his paper,
[Evidence-based cybersecurity policy? A meta-review of security control effectiveness](https://doi.org/10.1080/23738871.2024.2335461),
and his LinkedIn article, 
[What Works in Cybersecurity: MFA](https://www.linkedin.com/pulse/what-works-cybersecurity-mfa-daniel-woods-e7mke/).

In particular, research from [Microsoft](https://arxiv.org/abs/2305.00945) found that MFA reduced
the risk of compromise by 99%, and a similar study from
[Google](https://dl.acm.org/doi/abs/10.1145/3308558.3313481) reported that MFA "prevents 94% of
attacks rooted in phishing and 76% of targeted attacks", while cyber insurance claims data offer
conflicting evidence:
[Marsh (2023)](https://www.marsh.com/en/services/cyber-risk/insights/using-cybersecurity-analytics-to-prioritize-cybersecurity-investments.html)
found MFA to be the least effective control, and
[Marsh (2024)](https://www.guycarp.com/insights/2024/07/cyber-sleeper-threat-business-email-compromise.html)
found MFA to be the most effective control.

One aspect of MFA that Daniel discusses (briefly) is the idea that MFA deployment rate may explain
the contradictory results, as organizations may have a policy requiring MFA and still have gaps in
coverage, allowing credential attacks against accounts that don't have MFA enabled. This analysis
models how coverage affects the effectiveness of MFA using Monte Carlo simulation.

# Analysis

To model the impact of deployment rate on attack likelihood, we can simulate an attack against a
random user in the organization. If the user doesn't have MFA, the attack always succeeds (the base
rate). If the user has MFA, the attack has a 1% success rate (as I found in my
[MFA Risk Analysis](https://jabenninghoff.github.io/quantrr/example/analysis/mfa.html), a 99%
reduction in likelihood leads to dramatically lower expected losses).

## Simple Model

We can use the uniform distribution to determine if a user has MFA; if a randomly generated number
between 0 and 1 exceeds the MFA deployment rate, $r$, the user doesn't have MFA. (This will always
be true if $r = 0$ and never true if $r = 1$). For users with MFA, the attack succeeds if a second
random uniform number is less than or equal to 0.01.

```{r}
#| label: attack_succeeds
attack_succeeds <- function(deployment_rate) {
  if (runif(1) > deployment_rate) {
    # no MFA
    TRUE
  } else {
    # has MFA
    runif(1) <= 0.01
  }
}
```

Attacks with $r = 0$ should always succeed, and attacks with $r = 1$ should succeed about 1% of the
time:

```{r}
#| label: attack_success_test
attack_succeeds(0)
attack_succeeds(1)
sum(replicate(10000, attack_succeeds(1))) # should be close to 100
```

Instead of using `replicate()` for repeated trials, we can take advantage of R vectors, which are
faster (allowing more trials):

```{r}
#| label: attack_success_rate
attack_success_rate <- function(deployment_rate, runs = 1e5) {
  has_mfa <- runif(runs) <= deployment_rate
  attack_fails <- runif(runs) <= (has_mfa * 0.99)
  sum(!attack_fails) / runs
}
```

We should expect an attack success rate of 100% with 0% deployment and approximately 1% with 100%
deployment:

```{r}
#| label: attack_success_rate_test
attack_success_rate(0)
attack_success_rate(1)
```

## Effect of Deployment Rate

Model the effects of deployment rate on attack success using `purrr:map_dbl()`:

```{r}
#| label: deployment_success
deployment <- (1:100) / 100
success_rate <- map_dbl(deployment, attack_success_rate)
deployment_success <- data.frame(deployment, success_rate)
```

Plot the results:

```{r}
#| label: deployment_success_plot
#| fig-alt: "Plot of attacker success rate against MFA deployment rate, showing that attacker
#|          success rate decreases linearly as deployment rate increases."
deployment_success |>
  ggplot(aes(deployment, success_rate)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(title = "Effects of MFA deployment rate on attacker success rate") +
  labs(x = "MFA deployment rate", y = "Attacker success rate") +
  theme_quo()
```

## Economic Impact of Deployment Rate

Perhaps unsurprisingly in retrospect, there is a clear linear relationship between deployment rate
and attacker success rate. However, as the earlier
[MFA Analysis](https://jabenninghoff.github.io/quantrr/example/analysis/mfa.html) shows, expected
losses drop dramatically as the likelihood of an attack approaches 0. A plot from that analysis
compares baseline risk (no treatment) against a pessimistic estimate of MFA attack success rate
reduction (76%) and an optimistic reduction (99%):

```{r}
#| label: risk_le
#| fig-alt: "Plot showing likelihood of experiencing a loss exceeding a given dollar value by risk,
#|          comparing treatments."
load("objects/mfa-exports.RData")
risk_le
```

Interactive plot:

```{r}
#| label: risk_le_plotly
#| fig-alt: "Interactive plot showing likelihood of experiencing a loss exceeding a given dollar
#|          value by risk, comparing treatments."
ggplotly(risk_le + guides(color = "none"))
```

In the base case, the firm has a 3.2% chance of a loss exceeding $1M (about once every 30 years),
which drops to 0.80% with a 76% lower attack success rate (once every 125 years), and 0.04% with a
99% reduction (once every 2500 years).

## Successful Attack Frequency

Instead of expressing attacker success as a percentage (rate), we can instead express it in terms
of frequency; about how many attacks does it take before an attacker is successful?

```{r}
#| label: frequency_plot
#| fig-alt: "Plot of attacker success frequency against MFA deployment rate, showing that the number
#|          of attacks before success increases exponentially with deployment rate"
frequency_plot <- deployment_success |>
  mutate(
    deployment = percent(deployment, 0), success_rate = percent(success_rate),
    frequency = 1 / success_rate
  ) |>
  ggplot(aes(deployment, frequency, text = paste0("success_rate: ", success_rate))) +
  geom_point() +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(title = "Effects of MFA deployment rate on frequency of succesful attacks") +
  labs(x = "MFA deployment rate", y = "Frequency (attacker succeeds every n tries)") +
  theme_quo()

ggplotly(frequency_plot)
```

This plot more clearly explains the effect of deployment rate on how often attacks succeed; at 50%
deployment, the attack succeeds about every 2 tries. At a deployment rate of 91%, the attack
succeeds every 10 tries. A 100% deployment rate (every 100 days) is about twice as effective as a
99% deployment rate (every 50 days) - this is true on both plots, but easier to see with the second.

<!--
Also look at the effectiveness of control coverage vs defense-in-depth. How do two partially
deployed controls compare to a single widely deployed control?
-->

# Discussion

<!--
Discuss how this could be done with probability math, how use of Monte Carlo helps, importance of
intuitive visualizations in understanding risk (rate vs frequency).
-->
