---
title: "Risk Value Analysis"
author: "John Benninghoff"
date: '2024-11-03'
date-modified: '2024-11-03'
categories: risk quantification
order: 105
output:
  html_notebook:
    theme:
      version: 5
      preset: bootstrap
    css: assets/extra.css
    pandoc_args: --shift-heading-level-by=1
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

An exploration of the value of cybersecurity risk reduction.

# Questions/TODO

- [ ] Questions/TODO list here

```{r setup, message = FALSE, warning = FALSE}
library(poilog)
library(tibble)
library(dplyr)
library(quantrr)
```

# Background

What is the value of a cybersecurity program? Put another way, how much should an organization pay
to reduce the likelihood of a breach or the expected impact? In this analysis, we compare two firms,
one with typical breach rate and impact, and a second that makes investments to reduce their risk.
Using Monte Carlo simulation, we can calculate the value of this risk reduction.

# Baseline Risk

We can model baseline risk for a typical firm using
[quantrr](https://jabenninghoff.github.io/quantrr/) and data from the Cyentia 2022
[Information Risk Insights Study (IRIS)](https://www.cyentia.com/wp-content/uploads/IRIS-2022_Cyentia.pdf).

The 2022 IRIS found that the upper bound likelihood of a breach in the next year fit a Poisson
log-normal distribution with a mean ($\mu$) of -2.284585 and and standard deviation
($\sigma$) of 0.8690759.

As was done in the [breach rate analysis](breach-rates.Rmd), we can use trial and error to find a
reasonable value of $\lambda$ for a Poisson distribution that approximates these results:

```{r breach_table}
runs <- 1e6
lambda <- 0.138

breaches_poilog <- rpoilog(runs, mu = -2.284585, sig = 0.8690759, keep0 = TRUE)
breaches_pois <- rpois(runs, lambda = lambda)

breach_table <- function(breaches) {
  years <- length(breaches)
  tibble(
    "One or more" = sum(breaches >= 1) / years,
    "Two or more" = sum(breaches >= 2) / years,
    "Three or more" = sum(breaches >= 3) / years
  )
}

bind_rows(breach_table(breaches_poilog), breach_table(breaches_pois))
```

A Poisson distribution with a $\lambda$ of 0.138 approximates the Poisson log-normal model from the
Cyentia IRIS report.

```{r meanlog}
meanlog <- 12.55949
sdlog <- 3.068723
```

For the impact, we can use the log-normal loss model from IRIS, with a mean ($\mu$) of `r meanlog`
and standard deviation($\sigma$) of `r sdlog`.

Using the baseline parameters, we can simulate security events and losses over the next 10 years:

```{r baseline}
calc_risk("baseline", lambda, meanlog, sdlog, runs = 10)
```
