---
title: "Risk Quantification Demo"
author: "John Benninghoff"
date: '2024-08-15'
date-modified: '2024-08-15'
categories: risk quantification
order: 104
execute:
  echo: false
output:
  html_notebook:
    theme:
      version: 5
      preset: bootstrap
    css: assets/extra.css
    pandoc_args: --shift-heading-level-by=1
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

Risk Quantification demonstration for my SIRAcon 2024 talk, "[UnFAIR: Simplifying and Expanding Technology Risk Quantification](https://web.cvent.com/event/7f49b0a6-bca9-46fd-8245-a2deb671efee/websitePage:23d1376e-2723-411a-910a-0edf87b03015?session=956b9176-13ad-4dac-af3c-e8ccb30bae8a&shareLink=true)."

# Questions/TODO

- [ ] Add environment statement: describe the background and history of the system being analyzed, summary of existing state and description of the risks [builds trust and common ground through demonstrating shared understanding of the system, include a diagram if available]
- [ ] Add before and after: baseline risk vs after replacing system (investment), which allows calculation of ROI

```{r setup, message = FALSE, warning = FALSE}
library(readxl)
library(janitor)
library(validate)
library(dplyr)
library(formattable)
library(purrr)
library(ggplot2)
library(jbplot)
library(plotly)

# TODO: implement better solution
source("../R/risk.R") # nolint: undesirable_function_linter. quarto can't load library(security).
```

# Import

Import and validate data from Excel. The data in `demo.xlsx` is based on the examples developed
[here](rq-prototype.Rmd).

```{r import}
risks <- read_xlsx("data/demo.xlsx", sheet = "Risks") |>
  clean_names()

validate_risks <- local({
  validate_rules <- validator(
    risk_char = is.character(risk),
    risk_not_na = !is.na(risk),
    risk_not_blank = nchar(risk) > 0,
    desc_char = is.character(description),
    desc_not_na = !is.na(description),
    desc_not_blank = nchar(description) > 0
  )
  validate_out <- confront(risks, validate_rules)
  validate_summary <- summary(validate_out)

  # end processing on any failures, disallow NA
  stopifnot(
    all(validate_summary$fails == 0),
    all(validate_summary$nNA == 0),
    !all(validate_summary$error),
    !all(validate_summary$warning)
  )

  return(validate_out)
})

estimates <- read_xlsx("data/prototype.xlsx", sheet = "Estimates") |>
  clean_names() |>
  rename(
    lambda = frequency_per_yer, p05 = low_5_percent, p95 = high_95_percent, p50 = most_likely
  )

validate_estimates <- local({
  validate_rules <- validator(
    risk_not_na = !is.na(risk),
    risk_match = risk %in% risks$risk,
    expert_char = is.character(expert),
    expert_not_na = !is.na(expert),
    expert_not_blank = nchar(expert) > 0,
    lambda_num = is.numeric(lambda),
    lambda_pos = lambda > 0,
    p05_num = is.numeric(p05),
    p05_pos = p05 > 0,
    p95_num = is.numeric(p95),
    p95_pos = p95 > 0,
    p50_num = is.numeric(p50),
    p50_pos = p50 > 0
  )

  validate_out <- confront(estimates, validate_rules)
  validate_summary <- summary(validate_out)

  # end processing on any failures, allow NA
  stopifnot(
    all(validate_summary$fails == 0),
    !all(validate_summary$error),
    !all(validate_summary$warning)
  )

  return(validate_out)
})
```

# Environment Statement

(Describe the background and history of the system being analyzed, summary of existing state, and
extended description of the risks)

Risk descriptions:

```{r risks}
risks |>
  formattable(align = "l")
```

# Forecast

Forecast risk using Monte Carlo simulation. The average events and losses for each risk are
summarized below:

```{r forecast}
consensus <- estimates |>
  group_by(risk) |>
  summarize(across(lambda:p50, ~ mean(.x, na.rm = TRUE)))

consensus_params <- consensus |>
  mutate(as_tibble(lnorm_param(.data$p05, .data$p95, .data$p50)))

forecast <- consensus_params |>
  select(c("risk", "lambda", "meanlog", "sdlog")) |>
  pmap(calc_risk) |>
  list_rbind()

forecast |>
  group_by(risk) |>
  summarize(avg_events = mean(events), avg_losses = mean(losses)) |>
  mutate(avg_losses = currency(avg_losses, digits = 0L)) |>
  formattable(align = "l")
```

# Losses

Losses by risk separately and in aggregate:

```{r risk_hist, warning = FALSE}
forecast |>
  ggplot(aes(losses)) +
  facet_grid(vars(risk)) +
  geom_hist_bw(bins = 100) +
  scale_x_log10(labels = scales::label_currency(scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = NULL, title = "Losses by Risk") +
  theme_quo()

forecast |>
  ggplot(aes(losses, fill = risk)) +
  geom_histogram(bins = 100) +
  scale_x_log10(labels = scales::label_currency(scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = NULL, fill = "Risk", title = "Total Losses") +
  scale_fill_viridis_d() +
  theme_quo()
```

# Loss Exceedance Curves

Plot loss exceedance curves for all risks and combined risk.

```{r p95_max}
# calculate the largest p95 (95% quantile), to limit the x-axis for static plots
p95_max <- forecast |>
  group_by(risk) |>
  summarize(p95 = quantile(losses, 0.95)) |>
  pull(p95) |>
  max()
```

## By Risk

Plot loss exceedance curves for each risk:

```{r risk_le}
risk_le <- forecast |>
  ggplot(aes(losses)) +
  facet_grid(vars(risk)) +
  stat_ecdf(aes(y = after_stat(1 - y))) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_x_continuous(labels = scales::label_currency(scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = NULL, title = "Loss Exceedance by Risk")

risk_le +
  coord_cartesian(xlim = c(0, p95_max)) +
  theme_quo()
```

Interactive plot:

```{r risk_le_plotly}
ggplotly(risk_le + theme_minimal())
```

## Combined Risk

Plot loss exceedance curve for combined risk:

```{r combined_le}
combined_le <- forecast |>
  group_by(year) |>
  summarize(total_losses = sum(losses)) |>
  ggplot(aes(total_losses)) +
  stat_ecdf(aes(y = after_stat(1 - y))) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_x_continuous(labels = scales::label_currency(scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = NULL, title = "Combined Loss Exceedance")

combined_le +
  coord_cartesian(xlim = c(0, p95_max)) +
  theme_quo()
```

Interactive plot:

```{r combined_le_plotly}
ggplotly(combined_le + theme_minimal())
```

# Appendix

Additional details on the risk quantification analysis.

## Validation

Data validation results for Risks tab:

```{r validate_risks}
plot(validate_risks)
```

Data validation results for Estimates tab:

```{r validate_estimates}
plot(validate_estimates)
```

## Estimates

All risk estimates:

```{r estimates}
estimates |>
  mutate(across(p05:p50, ~ currency(.x, digits = 0L))) |>
  formattable(align = "l")
```

## Consensus Estimate

Using a simple average of all experts that provided an estimate (not blank/NA), this gives us a
consensus estimate for the three risks of:

```{r consensus}
consensus |>
  mutate(across(p05:p50, ~ currency(.x, digits = 0L))) |>
  formattable(align = "l")
```

The consensus estimates for p05 and p95 result in the following parameters for log-normal loss
magnitude. The p50 estimate is used to calculate the percentage difference from the actual median
(`mdiff`), a measure of estimate accuracy:

```{r consensus_params}
consensus_params |>
  mutate(across(p05:p50, ~ currency(.x, digits = 0L))) |>
  mutate(mdiff = percent(mdiff)) |>
  formattable(align = "l")
```

## Forecast Summary

A `summary()` of the forecast results.

```{r summary}
summary(forecast)
```
